---
title: "PCA of gene expression in CTCs from NSG-CDX-BR16 mice"
author: "Francesc Castro-Giner"
date: "2022-02-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  sce_dir: ./data/sce
---


## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"), fig.showtext = TRUE)

options(stringsAsFactors = FALSE)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(showtext)
library(scater)
library(scran)
library(PCAtools)
library(cowplot)
library(ggalt)
library(grid)
library(gridExtra)
library(knitr)
library(kableExtra)
```

Set font family for figures
```{r set-font}
font_add("Helvetica", "./configuration/fonts/Helvetica.ttc")
showtext_auto()
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Load color palettes
```{r color-palettes}
source("./configuration/rmarkdown/color_palettes.R")
```

Load functions
```{r load-functions}
source('./code/R-functions/pca_tools.r')
source('./code/R-functions/color_tools.r')
```

Load SingleCellExpression data
```{r load-sce}
use_sce <- readRDS(file.path(params$sce_dir, 'sce_br16.rds'))
```


## PCA analysis Data
Configuration
```{r pca-conf}
use_metavars <-   c(`Library size` = "sum", 
                    `Detected genes` = "detected", 
                    `Mitochondrial\nproportion` = "subsets_Mito_percent", 
                    `CTC type` = "sample_type", 
                    `Number of CTC` = "ctc_n", 
                    `Time point` = "timepoint",
                    `G1 score` = 'G1_score',
                    `G2M score` = 'G2M_score',
                    `S score` = 'S_score')
```

Quantify per-gene variation and select top-500 highly variable genes
```{r hvg-selection}
fit_res <- modelGeneVar(use_sce, assay.type = "logcpm")
fit_md <- metadata(fit_res)
hvg_var_bio <- getTopHVGs(fit_res, n=500)
use_genes <- hvg_var_bio
```

Generate PCA object and calculate elbow point. 
```{r pca-obj}
mat <- logcounts(use_sce)[use_genes,]
rownames(mat) <- rowData(use_sce[use_genes,])$gene_name
p <- PCAtools::pca(mat, metadata = colData(use_sce) %>% data.frame)
p$metadata$timepointf <- factor(p$metadata$timepoint, levels = c('active', 'resting'))
p$metadata$timepoint_sample_type_f <- factor(
  p$metadata$timepoint_sample_type,
  levels = c('active_ctc_single', 'active_ctc_cluster', 'active_ctc_cluster_wbc', 'resting_ctc_single', 'resting_ctc_cluster', 'resting_ctc_cluster_wbc'))
```

Calculate elbow point
```{r elbow-point}
elbow_point <- findElbowPoint(p$variance)
```

Add additional metadata to PCA object
```{r add-metavars}
p$metadata <- p$metadata %>% 
  mutate(`Library size` = sum, 
         `Detected genes` = detected, 
         `Mitochondrial\nproportion` = subsets_Mito_percent, 
         `CTC type` = sample_type, 
         `Number of CTC` = ctc_n, 
         `Time point` = timepoint,
         `G1 score` = G1_score,
         `G2M score` = G2M_score,
         `S score` = S_score)
```

## Correlation of PC eigenvectors with metadata
Heatmap showing the Pearson’s correlation coefficient of PC1-7 eigenvectors from gene expression with technical and biological variables in BR16-CDX CTCs. *P* values by two-sided Pearson’s correlation test (\**P* < 0.01, \*\**P* <0.001, \*\*\**P* <0.0001).

```{r br16_pca_eigencorplot, fig.width = 5.1}
use_cex <- 8/12
eigencorplot(
  p,
  components = getComponents(p, 1:elbow_point),
  metavars =  names(use_metavars),
  col = c( "blue2", "blue1", "black", "red1", "red2"),
  colCorval = 'white',
  scale = TRUE,
  main = 'PCs clinical correlations',
  plotRsquared = FALSE,
  signifSymbols = c("***", "**", "*", ""),
  signifCutpoints = c(0, 0.0001, 0.001, 0.01, 1),
  cexTitleX= use_cex,
  cexTitleY= use_cex,
  cexLabX = use_cex,
  cexLabY = use_cex,
  cexMain = use_cex,
  cexLabColKey = use_cex,
  cexCorval = use_cex
)
```

### Table : Percentage of variance associated to each PC
```{r pca_variance_tab}
p$variance[1:elbow_point] %>% data.frame %>% set_names('Variance') %>% 
  kable(caption = "Percentage of variance associated to each PC") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

### Table : Pearson r values
```{r pca_cor_r}
pca_cor_val <- pca_eigencorplot(p, components = getComponents(p, 1:elbow_point), metavars =  names(use_metavars), returnPlot = FALSE)
pca_cor_val$corvals %>% t %>% 
  kable(caption = "Pearson r values correlation values") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

### Table : Pearson correlation P-values
```{r pca_cor_p}
pca_cor_val$pvals_format <- apply(pca_cor_val$pvals, 2, format.pval, digits = 2)
dimnames(pca_cor_val$pvals_format) <- dimnames(pca_cor_val$pvals)
pca_cor_val$pvals_format %>% t %>% 
  kable(caption = "Pearson correlation P-values") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

## Biplot PC4 and PC5
Plot showing the principal components PC4 and PC5 of gene expression in CTCs from NSG-CDX-BR16 mice. *Upper* and *right* panels show the density of values for active (*blue*) and rest phase (*red*). 

```{r br16_pca_4_5_conf}
PCx <- 'PC4'
PCy<- 'PC5'

zt_sample_type_legend_palette_t <- sapply(zt_sample_type_legend_palette, transparent_col, percent = 50) 

use_palette <- c(zt_sample_type_legend_palette, timepoint_palette)
use_shapes <- c(
  'ZT16 Single CTCs' = 16,
  'ZT16 CTC-Clusters' = 17,
  'ZT16 CTC-WBC Clusters' = 15,
  'ZT4 Single CTCs' = 16,
  'ZT4 CTC-Clusters' = 17,
  'ZT4 CTC-WBC Clusters' = 15,
  active = 1,
  resting = 1
)
use_palette_sel <- c(
  'ZT16 Single CTCs' = use_palette['active'] %>% unname,
  'ZT16 CTC-Clusters' = use_palette['active'] %>% unname,
  'ZT16 CTC-WBC Clusters' = use_palette['active'] %>% unname,
  'ZT4 Single CTCs' = use_palette['resting'] %>% unname,
  'ZT4 CTC-Clusters' = use_palette['resting'] %>% unname,
  'ZT4 CTC-WBC Clusters' = use_palette['resting'] %>% unname,
  use_palette['active'],
  use_palette['resting']
  
)
circle_data <- cbind(p$metadata, 
                     x = p$rotated[,PCx],
                     y = p$rotated[,PCy])

xlab_name <- paste0(PCx,', ', p$variance[PCx] %>% round(2) %>% unname, '% variation')
ylab_name <- paste0(PCy,', ', p$variance[PCy] %>% round(2) %>% unname, '% variation')

biplot_res <- circle_data %>% 
  ggplot(aes(x, y)) +
  geom_point(
    aes(fill = zt_sample_type_legend, color = zt_sample_type_legend, shape = zt_sample_type_legend), 
    alpha = 0.4, size = 3
  ) +
  geom_encircle(
    aes(color = timepoint),
    alpha = 0.6,
    size = 1.5,
    s_shape = 1.5,
    show.legend = FALSE,
    na.rm = TRUE, 
    expand = 0) +
  scale_color_manual(values = use_palette_sel) +
  scale_fill_manual(values = use_palette_sel) +
  scale_shape_manual(values = use_shapes) +
  theme_cowplot(font_family = "Helvetica", font_size = 8, rel_small = 8/8, rel_tiny = 8/8, rel_large = 8/8)  +
  theme (
    axis.line = element_line(size = rel(0.25)),
    axis.ticks = element_line(size = rel(0.25)), 
    panel.border = element_rect(size = rel(1), fill = NA, colour = "black")
    ) +
  labs(
    x = xlab_name,
    y = ylab_name
  )
```

Main plot
```{r br16_pca_4_5, fig.asp = 0.9}
x_density_plot <- ggplot(circle_data, aes(x = x, fill = timepoint)) +
  geom_density(alpha = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = timepoint_palette) +
  theme (
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank()
    )


y_density_plot <- ggplot(circle_data, aes(x = y, fill = timepoint)) +
  geom_density(alpha = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = timepoint_palette) +
  theme (
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank()
    ) +
  coord_flip()


plot_grid(
  x_density_plot, NULL, NULL, 
  NULL, NULL, NULL, 
  biplot_res + theme(legend.position = "none"), NULL, y_density_plot,
  nrow = 3,
  ncol = 3,
  align="hv",
  axis = "tblr",
  rel_heights = c(1.1, -0.45, 3),
  rel_widths = c(3, -0.45, 1)
  )

```

Plot legend
```{r br16_pca_4_5_legend}
legend <- cowplot::get_legend(biplot_res)
grid.newpage()
grid.draw(legend)
```

